Implementing Distributed Systems Directly with I/O Automata in Ioa++

### 2011-09-22 papers/paper1/component_model.tex
Performed analysis again.

In this section we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
The Input/Output (I/O) automata model was developed by Lynch~\cite{lynch1996distributed} to model reactive concurrent and asynchronous systems.
An I/O automaton consists of state variables and a set of atomic actions that manipulate the state variables.
An action is only permitted to manipulate the state variables of the automaton to which it belongs.
An action is either an internal action, an output action, or an input action.
An internal action changes the state of the automaton.
An output action changes the state of the automaton and produces a signal or value that may be consumed by one or more input actions.
An input action changes the state of the automaton when it receives a signal or value produced by an output action.
\emph{Unvalued} output and input actions produce and consume signals respectively.
\emph{Valued} output and input actions produce and consume values respectively.
The set of actions associated with an automaton is known as the automaton's \emph{signature}~\cite{lynch1996distributed}.

@@

\begin{figure}
\center
\includegraphics[width=\textwidth]{system_model}
\caption{Example event detection system.}
\label{sys_model}
\end{figure}

@@

\paragraph{Recursive Encapsulation.}
Component frameworks that support recursive encapsulation allow components to contain other components.
Recursive encapsulation is at the heart of both bottom-up design strategies where smaller components are composed to produce larger ones, and top-down design strategies in which larger components are decomposed into smaller ones.
The contained component is the \emph{child} while the containing component is the \emph{parent}.
The resulting hierarchy helps developers organize programs by separating concerns and facilitates the reuse of existing components.
A distinction is made between the \emph{name} given to an instance and the \emph{type} of a child component to allow a parent component to contain multiple children of the same type.

@@

Figure~\ref{sys_model} shows a sample event detector system where automata are depicted as solid rectangles.
The type of the automaton appears in a box in the upper left or lower left corner, e.g., Processor.
The name of a named automaton appears after its type, e.g., Converter C.
Actions are depicted as dotted rectangles and bindings are indicate by dotted lines with an arrow pointing to the input action.
Output actions are suffixed with \emph{!}, e.g., \emph{detect!}, input actions are suffixed with \emph{?}, e.g., \emph{detected?}, and internal actions have neither suffix, e.g., \emph{process1}.
The type associated with a valued action is in parentheses after the name, e.g., \emph{out!(s)}.

@@

\paragraph{Explicit Composition.}
Composition in the I/O automata model is based on actions having the same name and can be applied to an arbitrary number of automata.
Using recursive encapsulation, developers must have the ability to compose automata regardless of the naming convention used by the original developers.
To prepare for dynamic composition, we limit the scope of composition to associating a single output action with a single input action.
Such an association is called a \emph{binding} and consists of a pair $(output, input)$ where $output$ is the name of the output action and $input$ is the name of the input action.

Each automaton contains a set of bindings referring to its own actions, the actions of its children, the actions of its children's children, etc.
The automaton that prescribes a binding is said to \emph{own} that binding.
The global set of bindings can be used to rename input actions before applying composition as defined by the I/O automata model.
A map for renaming inputs is generated by replacing action names with their fully-qualified names in all bindings by tracing from the root automaton.
For example, the fully-qualified name of the $in?(s)$ action of the Converter C in Figure~\ref{sys_model} is $.C.in?(s)$ due to the anonymous root automaton.
Composition as defined by the I/O automata model can be applied after using the map to rename all input actions to their corresponding output action.
The map of bindings for the event detection system is $\{ (.out!(s), .C.in?(s)), (.C.out!(t), .D.in?(t)), (.D.detect!, .detected?) \}$.

\paragraph{Binding rules.}
To be compliant with name-based composition under the I/O automata model, the set of bindings must adhere to certain rules.
First, the output action and input action of a binding must agree on the type being produced and consumed.
For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
The system shown in Figure~\ref{sys_model} follows this rule since $(out!(s), in?(s))$ agree on type $s$, $(out!(t), in?(t))$ agree on type $t$, and $(detect!, detected?)$ agree on the absence of a value (i.e., a signal).
Second, an input action can be bound to at most one output action.
Composing an input action with multiple output actions would introduce ambiguity into the relationship between the states of the automata and is therefore prohibited by the model.
A binding map where all bindings that mention the same input action are equivalent satisfies this rule.
For dynamics, which are discussed in Section~\ref{dynamics}, we must strengthen this rule to say that an input action can appear at most once in the global binding map.
This strengthened rule means that an input is the sink of at most one binding and that every binding has exactly one owner.
Third, an output action cannot be bound to more than one input action in the same automaton:  the state of the automaton containing the input actions would not be well-defined after the output is executed since the input action effects would be applied in an undefined order.
%% Let $\pi$ be a prefix operator that strips the action name (leaving the path and hence the automaton) of a fully-qualified action name.
%% The following must be true for the global binding map $B$:  $\langle \forall o : (o, z) \in B :: \langle \forall i,j : (o, i) \in B \land (o, j) \in B \land i \neq j:: \pi (i) \neq \pi (j) \rangle \rangle$.
Fourth, an output action cannot be bound to an input action in the same automaton.
Allowing such bindings would result in undefined behavior since the order of the output and input effects is undefined.
The I/O automata model prevents this by requiring that all actions belonging to the same automaton have a unique name.
To elaborate, we usually think about the output action occurring before the input action because a value must be generated.
However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
Graphically, an arrow originating in one component must terminate in another component, even if the latter is a child of the former as Figure~\ref{sys_model} illustrates.
%% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.

\paragraph{Concurrent execution.}
Recall that composing I/O automata consists of concatenating state vectors and folding input actions into output actions and that the result of composition is a single equivalent automaton.
Since the state of each automaton is independent, opportunities exist for concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
Each local action implies a set of automata which in turn implies a set of state variables that might be modified.
For an internal action, this set consists of the automaton containing the internal action.
For an output action, this set consists of the automaton containing the output action and the automata that contain the input actions to which the output action is bound.
Two actions can be executed concurrently if their respective sets of state variables are disjoint.
The implied automata sets for each local action in Figure~\ref{sys_model} are: $out!(s) \to \{root, C\}$, $out!(t) \to \{C, D\}$, $process1 \to \{D\}$, $process2 \to \{D\}$, and $detect! \to \{D, root\}$.
Thus, $process1$ and $out!(s)$ can be executed concurrently while $process1$ and $detect!$ can not.

@@

\paragraph{Parameters and action types.}
Actions in the I/O automata model can be defined using parameters.
Parameters are useful for situations requiring fan-in or session semantics.
For example, consider an automaton with an input action $in?(s)$ that is to be bound to $N$ different output actions.
To get around the limitation that an input can only be bound to one output, we augment the input action with a parameter that indicates the associated output action, e.g., $in?[i](s)$ where $i$ takes on the integer values from 1 to $N$.
\emph{Parameterized} actions take a parameter while \emph{unparameterized} actions do not.
Parameters, for actions that require them, are specified in the binding and serve to identify the action.

The allowed combinations of input, output, and internal actions, unvalued and valued actions, and unparameterized and parameterized actions result in 10 possible action types:
unvalued unparameterized output (uv-up output),
unvalued parameterized output (uv-p output),
valued unparameterized output (v-up output),
valued parameterized output (v-p output),
unvalued unparameterized input (uv-up input),
unvalued parameterized input (uv-p input),
valued unparameterized input (v-up input),
valued parameterized input (v-p input),
unparameterized internal (up internal), and
parameterized internal (p internal).

\begin{figure}
\center
\includegraphics[width=\textwidth]{example1}
\caption{Example remote sensor system.}
\label{example1}
\end{figure}

\paragraph{Example.}
Figure~\ref{example1} depicts a remote sensor system as a composition of automata using a scheme similar to Figure~\ref{sys_model} where parameters are given in brackets after the action name, e.g., \emph{recv?[0](m)}.
The Remote Sensor automaton consists of the Control automaton Z and three Channel automata A, B, and C representing three network connections.
The Control automaton Z consists of the Sensor automaton S, the Filter automaton F, and the Statistics automaton T.
The Control automaton C starts the sampling process with the uv-up output $Z.trigger!$.
The uv-up input $S.trigger?$ is executed atomically with $Z.trigger!$ due to the binding $(Z.trigger!, S.trigger?)$.
When the sampling process is complete, the v-up output $S.sample!(s)$ distributes the sample to the Filter F and Statistics component T via the $(S.sample!(s), F.in?(s))$ and $(S.sample!(s), T.in?(s))$ bindings.
Note that $S.sample!(s)$, $F.in?(s)$, and $T.in?(s)$ all agree on the type of the sample $s$.
The Filter automaton F passes the filtered sample to the Control automaton Z via the $(F.out!(t), Z.sample?(t))$ binding.
The Statistics automaton T calculates the statistics incrementally using the up internal $T.compute$.
The statistics can be polled by the Control automaton Z by issuing a request via the $(Z.request!, T.request?)$ binding and then receiving a response via the $(T.response!(u), Z.response?(u))$ binding.
The Control automaton Z contains a vp-output $Z.send![](m)$ and a vp-input $Z.recv?[](m)$ for sending and receiving network messages using the Channel automata A, B, and C.
The parameters 0, 1, and 2, are used to route messages to Channel automata A, B, and C respectively using the session idea mentioned previously.
Opportunities for concurrent execution exist for the remote sensor depicted in Figure~\ref{example1}.
The set of automata implied by $Z.trigger!$ is $\{Z, S\}$ while the set of automata implied by $T.compute$ is $\{T\}$.
Since the sets are disjoint, the two actions can be executed concurrently.

@@

The I/O automata model assumes that systems are composed of a finite or countably infinite number of automata~\cite{lynch1996distributed}.
Assuming an countably infinite number of participants allows properties of the systems to be stated in terms of an abstract quantity representing the number of members, e.g., $N$, and is reasonable because all real systems are composed of a finite number of members.
The set of interactions in the I/O automata model is also fixed since automata are statically composed.
Dynamic systems are modeled by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for ``waking up'' and ``going to sleep'' are defined~\cite{lynch1996distributed}.

@@

\paragraph{System actions.}
The four indivisible operations for managing a dynamic set of automata are \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
Create, bind, unbind, and destroy are collectively called \emph{system actions}.
The \emph{create} operation allows an automaton to create a child automaton.
The \emph{bind} operation allows an automaton to bind an input action to an output action while the \emph{unbind} operation allows an automaton to separate an input action from an output action.
The \emph{destroy} operation allows an automaton to recursively destroy a child automaton and all bindings associated with it.

@@

\begin{figure}
\center
\includegraphics[width=\textwidth]{system_action}
\caption{System actions.
  A base automaton contains output actions for creating, binding, unbinding, and destroying.
  The system automaton contains parameterized input actions for receiving the requests to create, binding, unbind, and destroy.
  The system automaton returns a result using the appropriate parameterized output action.
  The base automaton contains input actions to receive the results.}
\label{system_action}
\end{figure}

\paragraph{The system automaton.}
The set of automata $A$ and set of bindings $B$ are managed by the \emph{system automaton} which represents the run-time system.
Each automaton is composed (not bound) with the system automaton and has output actions for requesting system actions and input actions for receiving the results of system actions.
The system automaton, in turn, has input actions for receiving requests for system actions, possibly some internal actions for evaluating system action requests, and output actions for responding with the result of system actions.
Figure~\ref{system_action} shows the relationship between a base automaton and the system automaton.

@@

\paragraph{Binding predicates.}
The static configuration of a system as prescribed by the I/O automata model guarantees that the set of implied automata for an action is always the same.
This property is necessary and desirable for reasoning about the behavior of the system.
Admitting dynamics removes this guarantee since the set of implied automata will change according to the various system actions being executed.
To be specific, the set of implied automata for internal actions does not change but the set of implied automata for output actions can be changed with a successful bind, unbind, or destroy.

>>>>

In this section we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
The Input/Output (I/O) automata model was developed by Lynch to model reactive concurrent and asynchronous systems~\cite{lynch1996distributed}.
An I/O automaton consists of state variables and a set of atomic actions that manipulate the state variables.
An action is only permitted to manipulate the state variables of the automaton to which it belongs.
An action is either an internal action, an output action, or an input action.
An internal action changes the state of the automaton.
An output action changes the state of the automaton and produces a signal or value that may be consumed by one or more input actions.
An input action changes the state of the automaton when it receives a signal or value produced by an output action.
The set of actions associated with an automaton is known as the automaton's \emph{signature}~\cite{lynch1996distributed}.

@@

\begin{figure}
\center
\includegraphics[width=\columnwidth]{system_model}
\caption{Example event detection system.}
\label{sys_model}
\end{figure}

@@

\paragraph*{Recursive Encapsulation}
Component frameworks that support recursive encapsulation allow components to contain other components.
Recursive encapsulation is at the heart of both bottom-up design strategies where smaller components are composed to produce larger ones, and top-down design strategies in which larger components are decomposed into smaller ones.
The contained component is the \emph{child} while the containing component is the \emph{parent}.
The resulting hierarchy helps developers organize programs by separating concerns and facilitates the reuse of existing components.
A distinction is made between the \emph{name} given to an instance and the \emph{type} of a child component to allow a parent component to contain multiple children of the same type.

@@

Figure~\ref{sys_model} shows a sample event detector system where automata are depicted as solid rectangles.
The type of the automaton appears in a box in the upper left or lower left corner, e.g., Processor.
The name of a named automaton appears after its type, e.g., Converter C.
Actions are depicted as dotted rectangles and bindings are indicate by dotted lines with an arrow pointing to the input action.
Output actions are suffixed with \emph{!}, e.g., \emph{detect!}, input actions are suffixed with \emph{?}, e.g., \emph{detected?}, and internal actions have neither suffix, e.g., \emph{process1}.
The type associated with the value produced or consumed by an action is in parentheses after the name, e.g., \emph{out!(s)}.

@@

\paragraph*{Explicit Composition}
Composition in the I/O automata model is based on actions having the same name and can be applied to an arbitrary number of automata.
Using recursive encapsulation, developers must have the ability to compose automata regardless of the naming convention used by the original developers.
To prepare for dynamic composition, we limit the scope of composition to associating a single output action with a single input action.
Such an association is called a \emph{binding} and consists of a pair $(output, input)$ where $output$ is the name of the output action and $input$ is the name of the input action.

Each automaton contains a set of bindings referring to its own actions, the actions of its children, the actions of its children's children, etc.
The automaton that prescribes a binding is said to \emph{own} that binding.
The global set of bindings can be used to rename input actions before applying composition as defined by the I/O automata model.
A map for renaming inputs is generated by replacing action names with their fully-qualified names in all bindings by tracing from the root automaton.
For example, the fully-qualified name of the $in?(s)$ action of the Converter C in Figure~\ref{sys_model} is $[root].C.in?(s)$.
Composition as defined by the I/O automata model can be applied after using the map to rename all input actions to their corresponding output action.
The map of bindings for the event detection system is 
\begin{displaymath}
\begin{split}
\{ ([root].out!(s), [root].C.in?(s)),\\
   ([root].C.out!(t), [root].D.in?(t)),\\
   ([root].D.detect!, [root].detected?) \}
\end{split}
\end{displaymath}

\paragraph*{Binding rules}
To be compliant with name-based composition under the I/O automata model, the set of bindings must adhere to certain rules.
First, the output action and input action of a binding must agree on the type being produced and consumed.
For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
The system shown in Figure~\ref{sys_model} follows this rule since $(out!(s), in?(s))$ agree on type $s$, $(out!(t), in?(t))$ agree on type $t$, and $(detect!, detected?)$ agree on the absence of a value (i.e., a signal).
Second, an input action can be bound to at most one output action.
Composing an input action with multiple output actions would introduce ambiguity into the relationship between the states of the automata and is therefore prohibited by the model.
A binding map where all bindings that mention the same input action are equivalent satisfies this rule.
For dynamics, which are discussed in Section~\ref{dynamics}, we must strengthen this rule to say that an input action can appear at most once in the global binding map.
This strengthened rule means that an input is the sink of at most one binding and that every binding has exactly one owner.
Third, an output action cannot be bound to more than one input action in the same automaton:  the state of the automaton containing the input actions would not be well-defined after the output is executed since the input action effects would be applied in an undefined order.
%% Let $\pi$ be a prefix operator that strips the action name (leaving the path and hence the automaton) of a fully-qualified action name.
%% The following must be true for the global binding map $B$:  $\langle \forall o : (o, z) \in B :: \langle \forall i,j : (o, i) \in B \land (o, j) \in B \land i \neq j:: \pi (i) \neq \pi (j) \rangle \rangle$.
Fourth, an output action cannot be bound to an input action in the same automaton.
Allowing such bindings would result in undefined behavior since the order of the output and input effects is undefined.
The I/O automata model prevents this by requiring that all actions belonging to the same automaton have a unique name.
To elaborate, we usually think about the output action occurring before the input action because a value must be generated.
However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
Graphically, an arrow originating in one component must terminate in another component, even if the latter is a child of the former as Figure~\ref{sys_model} illustrates.
%% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.

\paragraph*{Concurrent execution}
Recall that composing I/O automata consists of concatenating state vectors and folding input actions into output actions and that the result of composition is a single equivalent automaton.
Since the state of each automaton is independent, opportunities exist for concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
Each local action implies a set of automata which in turn implies a set of state variables that might be modified.
For an internal action, this set consists of the automaton containing the internal action.
For an output action, this set consists of the automaton containing the output action and the automata that contain the input actions to which the output action is bound.
Two actions can be executed concurrently if their respective sets of state variables are disjoint.
The implied automata sets for each local action in Figure~\ref{sys_model} are: $out!(s) \to \{root, C\}$, $out!(t) \to \{C, D\}$, $process1 \to \{D\}$, $process2 \to \{D\}$, and $detect! \to \{D, root\}$.
Thus, $process1$ and $out!(s)$ can be executed concurrently while $process1$ and $detect!$ can not.

@@

\paragraph*{Parameters and action types}
Actions in the I/O automata model can be defined using parameters.
Parameters are useful for situations requiring fan-in or session semantics.
For example, consider an automaton with an input action $in?(s)$ that is to be bound to $N$ different output actions.
To get around the limitation that an input can only be bound to one output, we augment the input action with a parameter that indicates the associated output action, e.g., $in?[i](s)$ where $i$ takes on the integer values from 1 to $N$.
\emph{Parameterized} actions take a parameter while \emph{unparameterized} actions do not.
Parameters, for actions that require them, are specified in the binding and serve to identify the action.

\emph{Unvalued} output and input actions produce and consume signals respectively.
\emph{Valued} output and input actions produce and consume values respectively.
The allowed combinations of input, output, and internal actions, unvalued and valued actions, and unparameterized and parameterized actions result in 10 possible action types:
unvalued unparameterized output (uv-up output),
unvalued parameterized output (uv-p output),
valued unparameterized output (v-up output),
valued parameterized output (v-p output),
unvalued unparameterized input (uv-up input),
unvalued parameterized input (uv-p input),
valued unparameterized input (v-up input),
valued parameterized input (v-p input),
unparameterized internal (up internal), and
parameterized internal (p internal).

\begin{figure}
\center
\includegraphics[width=\columnwidth]{example1}
\caption{Example remote sensor system.}
\label{example1}
\end{figure}

\paragraph*{Example}
Figure~\ref{example1} depicts a remote sensor system as a composition of automata using a scheme similar to Figure~\ref{sys_model} where parameters are given in brackets after the action name, e.g., \emph{recv?[0](m)}.
The Remote Sensor automaton consists of the Control automaton Z and three Channel automata A, B, and C representing three network connections.
The Control automaton Z consists of the Sensor automaton S, the Filter automaton F, and the Statistics automaton T.
The Control automaton C starts the sampling process with the uv-up output $Z.trigger!$.
The uv-up input $S.trigger?$ is executed atomically with $Z.trigger!$ due to the binding $(Z.trigger!, S.trigger?)$.
When the sampling process is complete, the v-up output $S.sample!(s)$ distributes the sample to the Filter F and Statistics component T via the $(S.sample!(s), F.in?(s))$ and $(S.sample!(s), T.in?(s))$ bindings.
Note that $S.sample!(s)$, $F.in?(s)$, and $T.in?(s)$ all agree on the type of the sample $s$.
The Filter automaton F passes the filtered sample to the Control automaton Z via the $(F.out!(t), Z.sample?(t))$ binding.
The Statistics automaton T calculates the statistics incrementally using the up internal $T.compute$.
The statistics can be polled by the Control automaton Z by issuing a request via the $(Z.request!, T.request?)$ binding and then receiving a response via the $(T.response!(u), Z.response?(u))$ binding.
The Control automaton Z contains a vp-output $Z.send![](m)$ and a vp-input $Z.recv?[](m)$ for sending and receiving network messages using the Channel automata A, B, and C.
The parameters 0, 1, and 2, are used to route messages to Channel automata A, B, and C respectively using the session idea mentioned previously.
Opportunities for concurrent execution exist for the remote sensor depicted in Figure~\ref{example1}.
The set of automata implied by $Z.trigger!$ is $\{Z, S\}$ while the set of automata implied by $T.compute$ is $\{T\}$.
Since the sets are disjoint, the two actions can be executed concurrently.

@@

The I/O automata model assumes that systems are composed of a finite or countably infinite number of automata~\cite{lynch1996distributed}.
Assuming an countably infinite number of participants allows properties of the systems to be stated in terms of an abstract quantity representing the number of members, e.g., $N$, and is reasonable because all real systems are composed of a finite number of members.
The set of interactions in the I/O automata model is also fixed since automata are statically composed.
Dynamic systems are modeled by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for ``waking up'' and ``going to sleep'' are defined~\cite{lynch1996distributed}\footnote{Use the actual paper.  I think it has something to do with databases.}.

@@

\paragraph*{System actions}
The four indivisible operations for managing a dynamic set of automata are \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
Create, bind, unbind, and destroy are collectively called \emph{system actions}.
The \emph{create} operation allows an automaton to create a child automaton.
The \emph{bind} operation allows an automaton to bind an input action to an output action while the \emph{unbind} operation allows an automaton to separate an input action from an output action.
The \emph{destroy} operation allows an automaton to recursively destroy a child automaton and all bindings associated with it.

@@

\begin{figure}
\center
\includegraphics[width=\columnwidth]{system_action}
\caption{System actions.
  A base automaton contains output actions for creating, binding, unbinding, and destroying.
  The system automaton contains parameterized input actions for receiving the requests to create, binding, unbind, and destroy.
  The system automaton returns a result using the appropriate parameterized output action.
  The base automaton contains input actions to receive the results.}
\label{system_action}
\end{figure}

\paragraph*{The system automaton}
The set of automata $A$ and set of bindings $B$ are managed by the \emph{system automaton} which represents the run-time system.
Each automaton is composed (not bound) with the system automaton and has output actions for requesting system actions and input actions for receiving the results of system actions.
The system automaton, in turn, has input actions for receiving requests for system actions, internal actions for evaluating system action requests, and output actions for responding with the result of system actions.
Figure~\ref{system_action} shows the relationship between a base automaton and the system automaton.

@@

\paragraph*{Binding predicates}
The static configuration of a system as prescribed by the I/O automata model guarantees that the set of implied automata for an action is always the same.
This property is necessary and desirable for reasoning about the behavior of the system.
Admitting dynamics removes this guarantee since the set of implied automata will change according to the various system actions being executed.
To be specific, the set of implied automata for internal actions does not change but the set of implied automata for output actions can be changed with a successful bind, unbind, or destroy.

### 2011-09-22 papers/paper1/design.tex
Performed analysis again.

\begin{figure}
\center
\includegraphics[width=\textwidth]{architecture}
\caption{Framework architecture.}
\label{framework_architecture}
\end{figure}

@@

\paragraph{Explicit dynamic scheduling.}
The I/O automata model assumes that the scheduler knows about every local action in the system.
There are two approaches for achieving this in practice.
In the first approach, automata declare \emph{statically} all of their actions to the scheduler.
The controller then has the responsibility of cycling through the actions according to some fair policy.
Since the scheduler has no knowledge about what actions are enabled, i.e., have a precondition that is true, this approach reduces to trying each action with brute force.

@@

\paragraph{Binding predicates and scheduling.}
The preceding discussion on scheduling assumes that an action in one automaton cannot affect preconditions in another automaton.
This is not entirely correct when one considers the system automaton and binding predicates.
For example, imagine an automaton with a single output action whose only precondition is that it must be bound and assume that the automaton only schedules the action when enabled.
When the automaton is created, its only action is disabled.
Consequently, it will not schedule its action and therefore will not be executed.
Suppose that some time after the automaton is created, another automaton successfully binds to the output action.
The bind changes the precondition but the automaton will never be executed since it has no way of scheduling the output action.
While this example may seem contrived, it is representative of situations where an automaton is waiting for one or more of its outputs to be bound: a situation that occurs quite often in practice.
To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after a successful bind or unbind.

\paragraph{Actions of no-longer-existing automata.}
Every action in the set of actions considered by the controller is scheduled by its associated automaton.
Since automata can be destroyed, an action in the set of actions owned by the controller might belong to an automaton that no longer exists.
Our solution is to check if the automaton exists before executing an action.
This check is performed by the dispatcher and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
Our solution assigns each automaton a unique identifier called an \emph{automaton id (aid)} and that the set of actions in the scheduler turns over comparatively faster than the set of aids.

\paragraph{Delayed actions.}
Scheduling activities for a future time is a useful feature for a real system because it allows the system to sleep between activities.
The framework should permit the development of automata that act as timers and alarms.
The corresponding feature that must be supported by the framework is scheduling an action at a future time.
Conceptually, the action is held until the time associated with the action at which point it is added to the set of actions being considered by the scheduler.
Actions scheduled in the past or present are released immediately to the scheduler.
When the same action is scheduled at two different times, the earliest time is used.
% Things I didn't say:  No canceling.

\paragraph{File descriptor events.}
File descriptors are used to communicate with the outside world in a POSIX environment.
Observe that there is no concept of blocking in the I/O automata model and introducing blocking I/O operations into action effects adversely affects the asynchronous and concurrent nature of I/O automata.
Thus, we require techniques that allow a program (1) to wait until a file descriptor is ready for I/O and (2) to perform non-blocking I/O operations.
This implies the use of either the Proactor or Reactor pattern~\cite{schmidt2000pattern}.
For our framework, we used the Reactor pattern since (1) the Reactor pattern is simpler and (2) all major operating systems have non-blocking I/O and some kind of synchronous I/O multiplexing whereas asynchronous I/O is much less standardized.

@@

\paragraph{Global FIFO scheduler.}
The scheduler included with the framework is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics.
Scheduling is idempotent to prevent unbounded growth of the queue.
The global FIFO scheduler uses the Reactor pattern to implement delayed actions and file descriptor events.

@@

\paragraph{System action protocol.}
Our approach to system actions follows the model introduced in Section~\ref{dynamics}.
All automata are statically composed with a system automaton via system actions for dynamically creating child automata, binding external actions, unbinding external actions, and destroying child automata.
The system automaton implements a request-response protocol where an automaton requests a system action and the system automaton processes the request and responds with either success or failure.
Associated with each request is a \emph{subject} which is the automaton's local name for an automaton or binding.
Subjects allow an automaton to have multiple outstanding requests.

@@

\paragraph{Automaton and Binding managers.}
The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple applications but becomes unwieldy when creating complicated constellations.
The Automaton Managers and Binding Managers provide an individualized facade to the system action state machines.
Creating an Automaton Manager initiates the creation of a new child automaton and invoking a destroy method initiates the destruction of a child automaton.
Automaton Managers can be observed using the Observer~\cite{gamma1995design} pattern to determine when and how the creation/destruction process succeeds.
Binding Managers are similar except they do not begin the process of binding until the automata providing the output action and input action have been created, thus, Binding Managers observe Automaton Managers.

\paragraph{Generators.}
A \emph{generator} is a single-use factory for creating an instance of an automaton.
Generators are used to produce new automata as part of the create system action and are visible as arguments in the various layers of the framework.

@@

Ioa++ is implemented in C++ and available at \url{http://jrwilson.github.com/ioa/}.

>>>>

\begin{figure}
\center
\includegraphics[width=\columnwidth]{architecture}
\caption{Framework architecture.}
\label{framework_architecture}
\end{figure}

@@

\paragraph*{Explicit dynamic scheduling}
The I/O automata model assumes that the scheduler knows about every local action in the system.
There are two approaches for achieving this in practice.
In the first approach, automata declare \emph{statically} all of their actions to the scheduler.
The controller then has the responsibility of cycling through the actions according to some fair policy.
Since the scheduler has no knowledge about what actions are enabled, i.e., have a precondition that is true, this approach reduces to trying each action with brute force.

@@

\paragraph*{Binding predicates and scheduling}
The preceding discussion on scheduling assumes that an action in one automaton cannot affect preconditions in another automaton.
This is not entirely correct when one considers the system automaton and binding predicates.
For example, imagine an automaton with a single output action whose only precondition is that it must be bound and assume that the automaton only schedules the action when enabled.
When the automaton is created, its only action is disabled.
Consequently, it will not schedule its action and therefore will not be executed.
Suppose that some time after the automaton is created, another automaton successfully binds to the output action.
The bind changes the precondition but the automaton will never be executed since it has no way of scheduling the output action.
While this example may seem contrived, it is representative of situations where an automaton is waiting for one or more of its outputs to be bound: a situation that occurs quite often in practice.
To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after a successful bind or unbind.

\paragraph*{Actions of no-longer-existing automata}
Every action in the set of actions considered by the controller is scheduled by its associated automaton.
Since automata can be destroyed, an action in the set of actions owned by the controller might belong to an automaton that no longer exists.
Our solution is to check if the automaton exists before executing an action.
This check is performed by the dispatcher and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
Our solution assigns each automaton a unique identifier called an \emph{automaton id (aid)} and that the set of actions in the scheduler turns over comparatively faster than the set of aids.

\paragraph*{Delayed actions}
Scheduling activities for a future time is a useful feature for a real system because it allows the system to sleep between activities.
The framework should permit the development of automata that act as timers and alarms.
The corresponding feature that must be supported by the framework is scheduling an action at a future time.
Conceptually, the action is held until the time associated with the action at which point it is added to the set of actions being considered by the scheduler.
Actions scheduled in the past or present are released immediately to the scheduler.
When the same action is scheduled at two different times, the earliest time is used.
% Things I didn't say:  No canceling.

\paragraph*{File descriptor events}
File descriptors are used to communicate with the outside world in a POSIX environment.
Observe that there is no concept of blocking in the I/O automata model and introducing blocking I/O operations into action effects adversely affects the asynchronous and concurrent nature of I/O automata.
Thus, we require techniques that allow a program (1) to wait until a file descriptor is ready for I/O and (2) to perform non-blocking I/O operations.
This implies the use of either the Proactor or Reactor pattern~\cite{schmidt2000pattern}.
For our framework, we used the Reactor pattern since (1) the Reactor pattern is simpler and (2) all major operating systems have non-blocking I/O and some kind of synchronous I/O multiplexing whereas asynchronous I/O is much less standardized.

@@

\paragraph*{Global FIFO scheduler}
The scheduler included with the framework is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics.
Scheduling is idempotent to prevent unbounded growth of the queue.
The global FIFO scheduler uses the Reactor pattern to implement delayed actions and file descriptor events.

@@

\paragraph*{System action protocol}
Our approach to system actions follows the model introduced in Section~\ref{dynamics}.
All automata are statically composed with a system automaton via system actions for dynamically creating child automata, binding external actions, unbinding external actions, and destroying child automata.
The system automaton implements a request-response protocol where an automaton requests a system action and the system automaton processes the request and responds with either success or failure.
Associated with each request is a \emph{subject} which is the automaton's local name for an automaton or binding.
Subjects allow an automaton to have multiple outstanding requests.

@@

\paragraph*{Automaton and Binding managers}
The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple applications but becomes unwieldy when creating complicated constellations.
The Automaton Managers and Binding Managers provide an individualized facade to the system action state machines.
Creating an Automaton Manager initiates the creation of a new child automaton and invoking a destroy method initiates the destruction of a child automaton.
Automaton Managers can be observed using the Observer~\cite{gamma1995design} pattern to determine when and how the creation/destruction process succeeds.
Binding Managers are similar except they do not begin the process of binding until the automata providing the output action and input action have been created, thus, Binding Managers observe Automaton Managers.

\paragraph*{Generators}
A \emph{generator} is a single-use factory for creating an instance of an automaton.
Generators are used to produce new automata as part of the create system action and are visible as arguments in the various layers of the framework.

@@

Ioa++ is implemented in C++ for POSIX environments and is available at \url{http://jrwilson.github.com/ioa/}.
